---
title: "edxCapstone"
author: "Jose Arizpe"
date: "29 December 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Movielens project / edx Data Science capstone project

The following code follows the methodology used in the course for fitting the movielens data, based on producing predictions based on the following formula:

- Y~u,i~ = $\mu$ + b~i~ + b~u~ + $\varepsilon$~u,i~  

where  

- Y~u,i~ - is the prediction for movie i and user u  
- $\mu$ - is the mean of ratings  
- b~i~ - is the movie effect  
- b~u~ - is the user effect  
- $\varepsilon$~u,i~ - is the residual error after taking into account the effects above  

The method above has been enhanced with the following additions:  

1- Introduction of regularisation based on penalised least squares

2- Break-up of the movielens "genres" field into columns, one column per genre containing logical values, so this information will also be available to the fitting process

3- Fitting of the residuals using the augmented file (including the expanded genres) using the "rpart" tree-based model  

The highest improvement on RMSE is obtained when $\varepsilon$~u,i~, the residual error, is estimated on its own and apart from the other variables in the formula, using for this purpose the rpart tree-based model. The fitting of residuals is in turn greatly enhanced by the splitting of the "genres" field into individual logical columns for each genre encountered in the data.

The application of regularisation based on penalised least squares also improves RMSE but in a much more modest way than the fitting of residuals.

`The resulting RMSE (0.426187) can be seen at the bottom of the report.`

Here are the steps followed to obtain this result:

## Load and prepare data
  
  
**1- Load data using the code provided**  
  
Create edx and validation sets
. . . . . . .
  
  
```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
################################
# Create edx set, validation set
################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1, sample.kind="Rounding")
# if using R 3.5 or earlier, use `set.seed(1)` instead
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```
  
  
**2- Extract the year from the movie title**
```{r load-lubridate-library, include=FALSE}
library(lubridate)
```

```{r extract-year-train-set}
# CREATE DATE FIELD
edx <- edx %>% mutate(date = as_datetime(timestamp))
# EXTRACT YEAR AND TITLE FROM ORIGINAL “title” COLUMN
#    Extract the year into vector "y"
y <- str_match(edx$title, "^(.*)\\s\\((\\d{4})\\)$")[,3]
edx$year <- y ## Attach “y” as a new column to “edx”
#    Eliminate year data from title
movieTitles <- str_match(edx$title, "^(.*)\\s\\((\\d{4})\\)$")[,2]  
edx$title <- movieTitles ## Replace title with the stripped down new title
```
  
  
**3- Break-up the "Genres" column**
  
Into as many columns as there are defined genres with logical values for each movie highligthing the genres which those movies have been classsified against.  
```{r unbind-genres-train-set}
# FIND OUT ALL DIFFERENT GENRES THAT EXIST IN THE LIST OF MOVIES AND PUT THEM IN A
# TABLE CALLED allGenres
#   Create new data frame including only “movieId” and “genres”
movieGenres <- edx %>% select(movieId, genres) %>% unique() 
movieGenres <- suppressWarnings(movieGenres %>% separate("genres", c("g1", "g2", "g3",
                                                                     "g4", "g5", "g6",
                                                                     "g7", "g8", "g9",
                                                                     "g10"), "\\|"))
#   separate genres into different columns, assume no more than 10 genres per movie. 
#   Create 1-n table for “movieId” and “genre”. Use intermediate table t
t <- gather(movieGenres, "key", "genre", -movieId, na.rm = TRUE)  
allGenres <- t %>% select(genre) %>% unique() ## List all different genres found
#   Delete records with text in parenthesis, assuming these are not valid genre names
allGenres <- allGenres %>% filter(!str_detect(allGenres$genre, "^\\(.*\\)$"))
#   Create colums with genres
for (g in allGenres[,1]) { 
  edx <- cbind(edx, data.frame(nom = str_detect(edx$genres, g)))
  names(edx)[names(edx) == "nom"] <- str_replace(tolower(g), "-", "_")
}
```
(Note that this processing makes the assumption that no movie is identified with more than 10 genres. Also it assumes that genres in parentheses are not genre names but rather it means no genre has been identified.)

Here is a list of all genres found
```{r list-genres}
allGenres
```

And here is an extract of the extended "edx" data including the added columns
```{r show-new-edx}
as_tibble(head(edx))
```
  
  
**4- Repeat the previous steps for the "validation" set**  

. . . . . . .  

```{r extract-year-and-unbind-genres-test-set, echo=FALSE}
# DO THE SAME AS ABOVE FOR VALIDATION DATA

# CREATE DATE FIELD
validation <- validation %>% mutate(date = as_datetime(timestamp))

# EXTRACT YEAR AND TITLE FROM ORIGINAL “title” COLUMN
#    Extract the year into vector “y”
y <- str_match(validation$title, "^(.*)\\s\\((\\d{4})\\)$")[,3] 
validation$year <- y ## Attach “y” as a new column to “edx”
#    Eliminate year data from title
movieTitles <- str_match(validation$title, "^(.*)\\s\\((\\d{4})\\)$")[,2]
#    Replace title column with the stripped down new title
validation$title <- movieTitles  

# FIND OUT ALL DIFFERENT GENRES THAT EXIST IN THE LIST OF MOVIES AND PUT THEM IN A
# TABLE CALLED allGenres
#    Create new data frame including “movieId” and “genres”
movieGenres <- validation %>% select(movieId, genres) %>% unique() 
#    Separate genres into different columns
movieGenres <- suppressWarnings(movieGenres %>% separate("genres", c("g1", "g2", "g3",
                                                                     "g4", "g5", "g6",
                                                                     "g7", "g8", "g9",
                                                                     "g10"), "\\|"))
#    Create 1-n table for “movieId” and “genre”
t <- gather(movieGenres, "key", "genre", -movieId, na.rm = TRUE) 
allGenres <- t %>% select(genre) %>% unique() ## List all different genres found
#    Delete records with text in parenthesis
allGenres <- allGenres %>% filter(!str_detect(allGenres$genre, "^\\(.*\\)$")) 
#    Create colums with genres
for (g in allGenres[,1]) { 
  validation <- cbind(validation, data.frame(nom = str_detect(validation$genres, g)))
  names(validation)[names(validation) == "nom"] <- str_replace(tolower(g), "-", "_")
}
```

## Create the model / start fitting
  
  
**5- Naive approach**
  
Start with naive approach as initial benchmark to measure progress against:

```{r RMSE-function, echo=FALSE}
# Create RMSE function
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

```{r naive-RMSE}
# Naïve RSME
mu_hat <- mean(edx$rating)
naive_rmse <- RMSE(validation$rating, mu_hat)
```
```{r add-naive-RSME-results-to-table, echo=FALSE}
message("RMSE naive method = ", naive_rmse)
results <- data.frame(method = "Naive", rmse = naive_rmse)
#    Add entry in matrix
```
  
  
**6- Try rpart as overarching method for predicting ratings**
  
Predict ratings using a tree-based method due to the number of columns that have been generated. I have tried various methods but all of them end up running out of memory except "rpart". Here is the first shot at that:**

```{r rpart-ratings-fit}
set.seed(1)
library(rpart) # Fit ratings using “rpart”
fit <- rpart(rating~., data = edx)
predicted_ratings <- predict(fit, validation)
model_0_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r add-rpart-ratings-fit-results-to-table, echo=FALSE}
message("RMSE fit ratings using rpart = ", model_0_rmse)
results <- suppressWarnings(bind_rows(results, data_frame(method="rpart ratings fit",
                                                          rmse = model_0_rmse)))
```

As can be seen, initial results with rpart for directly predicting the end-ratings are not great and do not achieve many points, so go back to the method used in the course, using the following formula as explained above: 

(Y~u,i~ = $\mu$ + b~i~ + b~u~ + $\varepsilon$~u,i~)  
  
  
**7- Add movie effects (b~i~) to naive estimate ($\mu$):**
  
```{r movie-effects}
mu <- mean(edx$rating)
movie_avgs <- edx %>% group_by(movieId) %>% summarize(b_i = mean(rating - mu))
predicted_ratings <- mu + validation %>% left_join(movie_avgs, by='movieId') %>%
  pull(b_i)
model_1_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r add-movie-effects-results-to-table, echo=FALSE}
message("RMSE from adding movie effect to naive = ", model_1_rmse)
results <- suppressWarnings(bind_rows(results, data_frame(method="Naive + movie effects Model",
                                                          rmse = model_1_rmse)))
```
  
  
**8- Add user effects (b~u~) to naive estimate ($\mu$) + movie effects (b~i~):**
  
```{r movie-user-effects}
user_avgs <- edx %>% left_join(movie_avgs, by='movieId') %>% group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
predicted_ratings <- validation %>% left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>% mutate(pred = mu + b_i + b_u) %>% pull(pred)
model_2_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r add-movie-user-effects-to-table, echo=FALSE}
message("RMSE from adding movie effect and user effect to naive = ",
        model_2_rmse)
results <- suppressWarnings(bind_rows(results, 
                                      data_frame(method="Naive + movie & User effects Model", 
                                                 rmse = model_2_rmse)))
```
  
  
**9- Apply regularisation to the naive ($\mu$) + movie effects (b~i~) estimate**
  
Apply regularisation based on penalised least squares, and see if this improves the results. For this the data has to be partitioned in order to fin the lambda value that yields the lowest RMSE:  
```{r movie-effects-regularised}
# Movie effect with regularisation
# Partition data
set.seed(1)
test_index <- createDataPartition(edx$rating, times=1, p=0.9, list=FALSE)
edx_test <- edx[-test_index,]
edx_train <- edx[test_index,]
edx_test <- edx_test %>% 
  semi_join(edx_train, by = "movieId") %>%
  semi_join(edx_train, by = "userId")

# Determine the lambda
lambdas <- seq(0, 10, 0.25)
mu <- mean(edx_train$rating)
just_the_sum <- edx_train %>% group_by(movieId) %>% 
  summarize(s = sum(rating - mu), n_i = n())
rmses <- sapply(lambdas, function(l1){
  predicted_ratings <- edx_test %>% left_join(just_the_sum, by='movieId') %>% 
    mutate(b_i = s/(n_i+l1)) %>% mutate(pred = mu + b_i) %>% pull(pred)
  return(RMSE(predicted_ratings, edx_test$rating))
})
qplot(lambdas, rmses)  
l1 <- lambdas[which.min(rmses)]
# Calculate RMSE for better lambda
mu <- mean(edx$rating)
just_the_sum <- edx %>% group_by(movieId) %>% 
  summarize(s = sum(rating - mu), n_i = n())
predicted_ratings <- validation %>% 
  left_join(just_the_sum, by='movieId') %>% mutate(b_i = s/(n_i+l1)) %>%
  mutate(pred = mu + b_i) %>% pull(pred)
model_6_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r message-step-3, echo=FALSE}
message("RMSE from naive + regularised movie effect = ", model_6_rmse)
message("(Best lambda = ", l1, ")")
```

```{r add-movie-effects-regularised-results-to-table, echo=FALSE}
results <- suppressWarnings(bind_rows(results, 
                                      data_frame(method="Naive + regularised movie effects Model", rmse = model_6_rmse)))
```
  
  
**10- Apply regularisation to the naive ($\mu$) + movie effects (b~i~) + user effects (b~u~) estimate**
  
Apply regularisation based on penalised least squares, and see if this improves the results. For this the data has to be partitioned in order to fin the lambda value that yields the lowest RMSE:  
```{r movie-user-effects-regularised}
# Movie and user effect with regularisation
# Partition data
set.seed(1)
test_index <- createDataPartition(edx$rating, times=1, p=0.9, list=FALSE)
edx_test <- edx[-test_index,]
edx_train <- edx[test_index,]
edx_test <- edx_test %>% 
  semi_join(edx_train, by = "movieId") %>%
  semi_join(edx_train, by = "userId")

# Determine the lambda
lambdas <- seq(0, 10, 0.25)
rmses <- sapply(lambdas, function(l2){
  mu <- mean(edx_train$rating)
  b_i <- edx_train %>% group_by(movieId) %>% 
    summarize(b_i = sum(rating - mu)/(n()+l2))
  b_u <- edx_train %>% left_join(b_i, by="movieId") %>% 
    group_by(userId) %>% summarize(b_u = sum(rating - b_i - mu)/(n()+l2))
  predicted_ratings <- edx_test %>% left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>% mutate(pred = mu + b_i + b_u) %>% 
    pull(pred)
  return(RMSE(predicted_ratings, edx_test$rating))
})
qplot(lambdas, rmses)
l2 <- lambdas[which.min(rmses)]
# Calculate RMSE for better lambda
mu <- mean(edx$rating)
b_i <- edx %>% group_by(movieId) %>% 
  summarize(b_i = sum(rating - mu)/(n()+l2))
b_u <- edx %>% left_join(b_i, by="movieId") %>% 
  group_by(userId) %>% summarize(b_u = sum(rating - b_i - mu)/(n()+l2))
predicted_ratings <- validation %>% left_join(b_i, by = "movieId") %>%
  left_join(b_u, by = "userId") %>% mutate(pred = mu + b_i + b_u) %>% 
  pull(pred)
model_3_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r message-step-4, echo=FALSE}
message("RMSE from naive + regularised movie effect + user effect = ", model_3_rmse)
message("(Best lambda = ", l2, ")")
```
```{r add-movie-user-effects-regularised-results-to-table, echo=FALSE}
results <- suppressWarnings(bind_rows(results, 
                                      data_frame(method="Naive + regularised movie & user effects Model", rmse = model_3_rmse)))
```
  
  
**11- Model the residuals ($\varepsilon$~u,i~) using rpart**
  
Use the complete data breadth, initial columns + calculated year + additional genre columns created above.

As mentioned before, I have tried a variety of training methods but all of them run into memory problems except "rpart" which also yields an important decrease in RSME as can be seen below.

Code and RMSE results follow:  
```{r movie-user-effects-plus-rpart-residuals-fit}
# NOW MODEL RESIDUALS INDEPENDENTLY ON TOP OF THE PREVIOUS MODELS USING rpart ON THE 
# NEW GENRE COLUMNS CREATED BEFORE
# First identify the winning parameters for all calculations to follow
mu <- mean(edx$rating)
b_i <- edx %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu)/(n()+l2))
b_u <- edx %>% left_join(b_i, by="movieId") %>% group_by(userId) %>% 
  summarize(b_u = sum(rating - b_i - mu)/(n()+l2))
# Create residuals table with residuals on ratings made by users
residuals <- edx %>% left_join(b_i, by = "movieId") %>% 
  left_join(b_u, by = "userId") %>% mutate(residual = rating - mu - b_i - b_u) %>% 
  pull(residual)
edx <- bind_cols(edx, data.frame(residuals = residuals))
library(rpart) # Fit residuals using “rpart”
fit <- rpart(residuals~., data = edx)

# Calculate predicted ratings using mu + b_i + b_u + residuals, i.e. adding residuals 
# to previous best estimate
r_hat <- predict(fit, validation)
validation <- bind_cols(validation, data.frame(residuals = r_hat))

predicted_ratings <- validation %>% left_join(b_i, by = "movieId") %>% 
  left_join(b_u, by = "userId") %>% mutate(pred = mu + b_i + b_u + residuals) %>% 
  pull(pred)
model_4_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r add-movie-user-effects-plus-rpart-residuals-fit-results-to-table, echo=FALSE}
message("RMSE from naive + regularised movie & user effects + rpart residuals fit = ",
        model_4_rmse)
results <- suppressWarnings(bind_rows(results, 
                                      data_frame(method="Naive + regularised movie & user effects + rpart residuals fit", rmse = model_4_rmse)))
```
  
  
**12- Model the residuals ($\varepsilon$~u,i~) without previous regularisation**
  
Just to be sure, and taking into account that the RMSE improves very little with Regularisation, model the residuals same as in the previus step but using the non-regularised naive + movie + user effects results.  
```{r movie-user-effects-plus-rpart-residuals-fit-witout-regularisation}
# NOW MODEL RESIDUALS SAME AS ABOVE BUT WITHOUT REGULARISATION IN THE PREVIOUS STEP 
# First identify the winning parameters for all calculations to follow
mu <- mean(edx$rating)
b_i <- edx %>% group_by(movieId) %>% summarize(b_i = mean(rating - mu))
b_u <- edx %>% left_join(b_i, by="movieId") %>% group_by(userId) %>% 
  summarize(b_u = mean(rating - b_i - mu))
# Create residuals table with residuals on ratings made by users
residuals <- edx %>% left_join(b_i, by = "movieId") %>% 
  left_join(b_u, by = "userId") %>% mutate(residual = rating - mu - b_i - b_u) %>% 
  pull(residual)
edx <- edx[ ,-which(names(edx)=="residuals")]
edx <- bind_cols(edx, data.frame(residuals = residuals))
fit <- rpart(residuals~., data = edx) # Fit residuals using “rpart”

# Calcuate predicted ratings using mu + b_i + b_u + residuals, i.e. adding residuals 
# to previous best estimate
r_hat <- predict(fit, validation)
validation <- validation[ ,-which(names(validation)=="residuals")]
validation <- bind_cols(validation, data.frame(residuals = r_hat))

predicted_ratings <- validation %>% left_join(b_i, by = "movieId") %>% 
  left_join(b_u, by = "userId") %>% mutate(pred = mu + b_i + b_u + residuals) %>% 
  pull(pred)
model_5_rmse <- RMSE(predicted_ratings, validation$rating)
```
```{r add-movie-user-effects-plus-rpart-residuals-wo-reg-fit-results-to-table, echo=FALSE}
message("RMSE from naive + movie & user effects + rpart residuals fit = ",
        model_5_rmse)
results <- suppressWarnings(bind_rows(results, 
                                      data_frame(method="Naive + movie & user effects + rpart residuals fit", rmse = model_5_rmse)))
```

## Show the results
  
  
**13- Results summary:**  

```{r results-table-print, echo=FALSE}
results %>% arrange(rmse)


message("Final RMSE = ", min(results$rmse))
message("delivered by '", results$method[which.min(results$rmse)], "'")
```
